---
title: "hyperSpec.tidyverse"
author: "Claudia Beleites"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{hyperSpec.tidyverse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{hyperSpec,magrittr}
---

```{r setup, include = FALSE}
library(hyperSpec.tidyverse)
library(hyperSpec)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  results = "hide"
)
```

This vignette introduces package `hyperSpec.tidytverse`.  
`hyperSpec.tidyverse` "fortifies" `hyperSpec` objects so that they can be used in `tidyverse` style.
In particular, this [piping](https://magrittr.tidyverse.org/) becomes convenient. 
On a more technical level, `hyperSpec.tidytverse` provides functions such as `filter`, `select`, and so on for `hyperSpec` objects.

## The Spectra Matrix

The spectra matrix in `$spc` behaves a bit different from the extra-data columns, because it contains a whole matrix inside one column.
Many `tidyverse` functions can directly deal with this.
In other cases, special attention is needed - 
this is explained in this vignette. 

## Selecting a subset of spectra with `filter()` or `slice()`

[dplyr::filter()](https://dplyr.tidyverse.org/reference/filter.html) selects rows, and `filter`ing by extra data columns works as with any `data.frame`:

```{r filter-extra-data}
flu %>% 
  filter (c > 0.2)

flu %>% 
  filter (c %>% between (0.15, 0.25))
```

The filter conditions (logical predicates) must yield one logical value per row. 
Logical expressions on the spectra matrix such as `spc > 100` yield a logical *matrix*.
This matrix indicates which *elements* of the spectra matrix match the condition.

To obtain the logical vector that indicates which *spectra* (rows) should be kept, further information is needed:
should a spectrum be kept if *any* of its intensity values (spectra matrix elements) fulfills the condition, 
or only if *all* of them do?
`hyperSpec` functions `any_wl()` and `all_wl()` summarize the logical matrix in this way.

Typical filtering tasks based on the spectra matrix are:

- remove "empty" spectra which have `NA`s at all wavelengths (which may be produced by certain import functions):
  ```{r filter spc empty spectra}
  flu %>%
    filter (!all_wl (is.na (spc)))
  ```
- remove spectra which have `NA`s at any wavelength (for example after spikes or other artifacts have been marked by `NA`s):
  ```{r filter spc NAs}
  flu %>%
    filter (!any_wl (is.na (spc)))
  ```
- keep only spectra with all intensities inside a specified range (e.g. non-negative and below a saturation threshold):
  ```{r filter spc between}
  flu %>%
    filter (all_wl (between (spc, 0, 500)))
  ```
- keep only spectra with *average* intensity inside a specified range (e.g. non-negative and below a saturation threshold):
  ```{r filter spc avg intensity between}
  flu %>% 
    filter (rowMeans (spc) %>% between (0, 500))
  ```
  Here, `any_wl()` or `all_wl()` are not needed: the spectra are already summarized into a single intensity value by `rowMeans()`.
  The condition therefore evaluates to a one logical value per spectrum already - which is what is needed.

To select rows by indices rather than a condition, use [slice()](https://dplyr.tidyverse.org/reference/slice.html):

```{r slice}
chondro %>% 
  slice (1:3)

chondro %>% 
  slice (800 : n())

chondro %>% 
  slice (-10 : -n())

flu %>% 
  slice (1, 3, 5)
```

Also, `head()` and `tail()` work as usual.

## Selecting particular data columns: `select()`

`select()` selects or discards particular columns from a `hyperSpec` object. 
If the spectra matrix (colum `spc`) is included in the selection, the result is still a `hyperSpec` object:

```{r, results='show'}
chondro %>% 
  select (clusters, spc)
```

If the result does not have the spectra matrix *in `$spc`*, it will be returned as `data.frame`:

```{r, results='show'}
flu %>% 
  select (-spc) 
```

To convert such a `data.frame` into a `hyperSpec` object again, use `as.hyperSpec`:

```{r, results='show'}
flu %>% 
  select (-spc) %>%
  as.hyperSpec ()
```

The resulting `hyperSpec` object has 0 wavelengths: its `$spc` column contains a spectra matrix with 0 columns, and its wavelength vector is also of length 0.
Behind the scenes, the `data.frame` returned by `select()` gets an attribute `labels` which stores the labels of the `hyperSpec` object.
`as.hyperSpec()` restores the labels from this attribute (if available).
Therfore, the "back-converted" `hyperSpec` object has its labels preserved.

## Selecting wavelength ranges

TODO


---
```{r, results='show'}
sessionInfo()
```

