---
title: "hyperSpec.tidyverse"
author: "Claudia Beleites"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{hyperSpec.tidyverse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library (hyperSpec.tidyverse)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  results = "hide"
)
```

This vignette gives a short introduction into package `hyperSpec.tidyverse` which "fortifies" `hyperSpec` objects for the use with `tidyverse` functions, i.e. it provides functions such as `filter`, `select`, and so on for `hyperSpec` objects.

This includes working with pipes (see [magrittr::magrittr]).

## The Spectra Matrix

The spectra matrix in `$spc` behaves a bit different from the extra-data columns, because it contains a whole matrix inside one column.

## Selecting a subset of spectra with `filter()`

[dplyr::filter()](https://dplyr.tidyverse.org/reference/filter.html) selects rows, and `filter`ing by extra data columns works as with any `data.frame`:

```{r filter-extra-data}
flu %>% 
  filter (c > 0.2)

flu %>% 
  filter (c %>% between (0.15, 0.25))
```

The filter conditions (logical predicates) need to evaluate to one logical value per row of the object that is `filter`ed. 
Logical expressions on the spectra matrix such as `spc > 100` produce a logical *matrix*, though, that indicates which *elements* of the spectra matrix match the condition. 
To obtain the logical *vector* that indicates which spectra (rows) should be kept, such a logical matrix can be converted into the needed vector by `hyperSpec::any_wl()` or `hyperSpec::all_wl()` which are `TRUE` for the spectra that fulfill the condition at any or all wavelengths, respectively.

Typical filtering tasks based on the spectra matrix are:

- remove "empty" spectra which have `NA`s at all wavelengths (which may be produced by certain import functions):
  ```{r filter spc empty spectra}
  flu %>%
    filter (!all_wl (is.na (spc)))
  ```
- remove spectra which have `NA`s at any wavelength (for example after spikes or other artifacts have been marked by `NA`s):
  ```{r filter spc NAs}
  flu %>%
    filter (!any_wl (is.na (spc)))
  ```
- keep only spectra with all intensities inside a specified range (e.g. non-negative and below a saturation threshold):
  ```{r filter spc between}
  flu %>%
    filter (all_wl (between (spc, 0, 500)))
  ```
- keep only spectra with average intensity inside a specified range (e.g. non-negative and below a saturation threshold):
  ```{r filter spc avg intensity between}
  flu %>% 
    filter (rowMeans (spc) %>% between (0, 500))
  ```

If the rows should be selected by indices rather than a condition, use [slice()](https://dplyr.tidyverse.org/reference/slice.html):

```{r slice}
  chondro %>% slice (1:3)
  chondro %>% slice (800 : n())
  chondro %>% slice (-10 : -n())
  flu %>% slice (1, 3, 5)
```

Also, `head` and `tail` work as usual.

## Select particular columns

[dplyr::select()](https://dplyr.tidyverse.org/reference/select.html)
